\chameleon{}: Untangling Type Errors Through Interactive Visualization and Exploration

### 2022-12-08 main.tex
A temporary save
More Tim's edits

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{tabularx}
\usepackage{minted}
\usepackage{graphicx}
\usepackage[hidelinks]{hyperref}

@@

@@

\begin{abstract}
In recent years, many programming communities have been undergoing a transition to writing strongly typed programs. To benefit from type-safe codebases, programmers must first overcome the challenge of resolving type errors where text messages can often be unhelpful and misleading. Modern programming tools do not concentrate on making type errors easy to understand. We demonstrate \chameleon{}, a type debugging tool capable of displaying the full context of a type error and allowing programmers to search for where the type error actually occurs based on their prior knowledge and to incrementally verify their assumptions. \chameleon{} was developed through a user-centered, iterative process with four participatory experiments. Through the experiments, we showed that programmers using \chameleon{} fix type errors faster than traditional text-based error messages.  This difference is more significant in harder tasks. Further, programmers actively using \chameleon{} interactive features are shown to be more efficient in fixing type errors than simply reading the type error output.
\end{abstract}

\begin{IEEEkeywords}
component, formatting, style, styling, insert
\end{IEEEkeywords}

\input{sections/introduction.tex}
\input{sections/related-work.tex}
\input{sections/chameleon.tex}
\input{sections/walkthrough.tex}
\input{sections/evaluation.tex}
\input{sections/discussion.tex}
\input{sections/conclusion.tex}

\bibliographystyle{IEEEtran}
\bibliography{icpc}

>>>>

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{tabularx}
\usepackage{minted}
\makeatletter
\let\@float@c@listing\@caption
\makeatother
\usepackage{graphicx}
\usepackage[hidelinks]{hyperref}

@@

% Comment out second line to disable.
\newcommand{\todo}[1]{}
% \renewcommand{\todo}[1]{{\color{red} TODO: {#1}}}

@@

\begin{abstract}

% \todo{I would start with saying untyped langs popular e.g. Python and the need to address problems this brings...}

Dynamically-typed programming languages are popular in education and the software industry. While presenting a low barrier to entry, they suffer from run-time type errors and longer-term problems in code quality and maintainability. Statically typed languages, while showing strength in these aspects, lack in learnability and ease of use. In particular, fixing type errors poses challenges to both novice users and experts. Further, compiler type error messages are presented in a static way that is biased towards the first occurence of the error in the program code. To help users resolve such type errors we introduce \chameleon{}, a type debugging tool that presents type errors to the user in an unbiased way, allowing them to explore the full context of where the errors could occur. Programmers can interactively verify the steps of reasoning against their intention. Through three studies involving real programmers, we showed that  \chameleon{} is more effective in fixing type errors than traditional text-based error messages. This difference is more significant in harder tasks. Further, programmers actively using \chameleon{}'s interactive features are shown to be more efficient in fixing type errors than passively reading the type error output.
\end{abstract}

\begin{IEEEkeywords}
types, type errors, debugging, visualization, exploration
\end{IEEEkeywords}

\input{sections/introduction.tex}
\input{sections/chameleon.tex}
\input{sections/walkthrough.tex}
\input{sections/evaluation.tex}
\input{sections/discussion.tex}
\input{sections/related-work.tex}
\input{sections/conclusion.tex}

\bibliographystyle{IEEEtran}
\bibliography{ICPC}

### 2022-12-08 sections/chameleon.tex
More Tim's edits

\section{Chameleon IDE} \label{chameleon}

\chameleon{} comprises two parts: a type inference engine and an original debugging interface. This separate architecture allows us to 1) adapt \chameleon{} into multiple front ends, such as IDE extensions, desktop applications, and web applications, and 2) reuse the debugging interface for different programming languages by replacing the back-end type inference engine. The debugging interface is designed from the ground up; the type inference engine is a re-implementation of the original Chameleon with several novel improvements.

@@

Chameleon was originally a command-line tool developed in the early 2000s to improve type error reporting and introduced a design for interactive type debugging for the Haskell programming language. Unlike traditional type errors produced by the Glasgow Haskell Compiler (GHC)~\cite{ben_gamari_home_2022}, which uses a Hindley–Milner type inference system, Chameleon infers types using constraint solving. In Chameleon, Constraints are generated from the source code based on typing rules. In addition, each constraint is labeled with the location (line and column numbers) where it is generated. This set of constraints is consistent if the program is well-typed and inconsistent otherwise. When a type error occurs, an efficient algorithm is used to derive a minimal subset of the constraints that still contain inconsistencies. This subset is called a Minimal Unsatisfiable Subset (MUS). From this, Chameleon can report a list of locations, using the labels of constraints that are in the MUS. Stuckey et al.~\cite{stuckey_interactive_2003} showed that program locations linked to the constraints from a MUS are all relevant to the type error and must include the cause of the error. In this paper, we will refer to these locations as \textit{error locations}.

Despite successfully borrowing the underlying ideas, we could not reuse the implementation of Chameleon since the project language standard and libraries used were out of date. In addition to what Chameleon can do, a few new type inference features were added in the \chameleon{} implementation.

Using only constraints from the MUS is sufficient to locate the type error, but to recover types from type errors we need  constraints from parts of the program that are irrelevant to the type error.  For instance, consider an ill-typed 2-tuple where two possible types can be assigned: \texttt{(Int, Int)} and \texttt{(Int, String)}. The types reconstructed from Chameleon may be \texttt{(a, Int)} and \texttt{(a, String)}. Although the recovered types are theoretically correct, they introduce a new type variable \texttt{a} making the error message harder to understand.  To solve this issue in \chameleon{}, we first expand the MUS to a maximally satisfiable subset. We do so by removing one constraint from the MUS to get a consistent set, and then repeatedly adding constraints from the global constraint store while maintaining consistency. This process stops when no more constraints can be added. The resulting subset, while not helpful to inform error locations, will produce the most concrete conflicting types. This difference can be seen in figure~\ref{fig:compare-to-original}.

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/compare-to-original.pdf}
    \caption{
Reporting the same type error, Chameleon uses
\texttt{Int -> a} and \texttt{Char -> a}, while \chameleon{} uses the 
concrete types \texttt{Int -> Bool} and \texttt{Char -> Bool}.
    }
    \label{fig:compare-to-original}
\end{figure}

In addition, \chameleon{} provides support for type explanation. Similar to the type explanation system ~\cite{jun_explaining_2002} from Yang,  \chameleon{} is able to produce a human-readable explanation, but for type errors. This is achieved by expanding the metadata fields of each constraint to contain the information needed for type explanation.

% \begin{lstlisting} [
%     language=Haskell, caption={
%     A simple program that is ill-typed. This program generates two constraints from line 1 and one constraint from line 2.
% }, label={listing:ifelse}
% ]
% if a then b else c
% a = "True"
% \end{lstlisting}
\begin{listing}[!ht]
\begin{minted}{haskell}
if a then b else c
a = "True"
\end{minted}
\caption{A simple program that is ill-typed. This program generates two constraints from line 1 and one constraint from line 2.}
\label{listing:1}
\end{listing}
For instance, the program in the listing \ref{listing:1} \chameleon{}  generates the following constraints and labels (in brackets) $T_a = Bool$ (if condition), $T_b = T_c$ (if branches), $T_a= String$  (definition). Clearly, as $T_a$ can not unify with both \textit{Bool} and \textit{String}, this program is not well typed. \chameleon{} can construct a human-readable explanation from the MUS, for example, \texttt{a} has type \texttt{Bool} because \texttt{a} is the condition of an if statement; however, \texttt{a} has type \texttt{String} because \texttt{a} is defined as the string literal \texttt{"True"}. This low-level explanation is used in a high-level user interface described in section \ref{sub:deduction-steps} to help programmers reason about the validity of error locations.

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{images/atonomy.pdf}
    \caption{
        \textbf{The anatomy of \chameleon{}.}
        \chameleon{} consists of two panes: an editor pane on the left and a
        debugging pane on the right. The editor pane is similar to a
        traditional code editor. Fragments of source code may have a highlight
        color (A). Additionally, an explanation layer (B) displays in the editor
        window, if deduction steps are enabled. The debugging window contains three blocks. 
        An error statement and configuration block, containing
        An error statement (D), and, optionally,
        a list of candidate expression cards (E), a list of deduction steps (F), 
        and a control bar (G) to increment/decrement deduction step. 
        The conflicting types
        block shows two alternative types (H) inferred by the type inference
        engine. A relevant type information block to show additional information (I)
        that may help understand type errors.
    }
    \label{fig:anatomy}
\end{figure*}

@@

The \chameleon{} debugging interface provides three main features to visualize the relation between an error statement and the error locations in the source code and to explore the explanation of each error location given by the type inference engine. Further, programmers  can turn on and off features to suit their preferences and debugging needs.

\paragraph{Type compare tool} \label{sub:type-compare}
In designing \chameleon{}, we want to argue against the practice of reporting type errors as the conflicts between \texttt{Expected type} and \texttt{Actual type}. Technically, the two alternatives are merely the side effect of different unification orders of the internal type-checking technique and have no bearing on what the programmers think and expect.

The type compare tool shows the two conflicting types in different colors, each associated with one or more error locations (Fig. \ref{fig:compare}).  The type error must be fixed by modifying at least one of these locations. The error locations are highlighted with the matching color of the conflicting types. The type compare tool is useful to quickly bisect the type error. If the programmers know the expression's intended type (they usually do), they will be able to eliminate half of the possible locations. To make the bisecting effect more pronounced, a user can hover over one of the possible types and only show the relevant locations that
contribute to that type. This is a convenient way to put the error \texttt{under the spotlight}.

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/intro-compare.pdf}
    \caption{
        \textbf{\chameleon{} minimal output (with candidate expression cards and deduction step disabled)}. \chameleon{} identified the conflicting types for the function \texttt{a} and associated the relevant locations with each type. In addition, \chameleon{} leaves out the right-hand side of the declaration '\texttt{0}' and '\texttt{1}' because they are unrelated.
}
    \label{fig:compare}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{images/intro-candidate.pdf}
    \caption{
        \textbf{\chameleon{} with only candidate expression cards enabled.}
        \chameleon{} shows that the type error can
        occur in the definition of '\texttt{x}' or '\texttt{y}'.
    }
    \label{fig:expression}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{images/intro-deduction.pdf}
    \caption{
        \textbf{\chameleon{} with both candidate expression cards and deduction steps
        enabled.}
        \chameleon{} explains the type error in four steps. In the screenshot, the active step is step 2, where \chameleon{} shows that the expression \texttt{x} and \texttt{y} should have the same type. 
    }
    \label{fig:deduction}
\end{figure}

@@

A candidate expression is an expression that can be inferred to have two conflicting types. 
We propose candidate expression cards to make it clear to the user that there are multiple ways to resolve a type conflict. By contrast,  standard compiler error messages such as those of GHC arbitrarily focus the users' attention on only one candidate expression, e.g.\ ``\textit{Couldn't match expected type ‘Char’ with actual type ‘Bool’ in expression x}" where x is the candidate expression. In practice, this candidate expression often does not match programmers' intention (Fig. \ref{fig:single-candidate}).

When a type error is detected, \chameleon{} provides a list of all candidate expressions and programmers are free to choose the problem to resolve by clicking on one candidate expression card. In the example shown in Fig. \ref{fig:expression}, \texttt{x} and \texttt{y} are both candidate expressions. Indeed, fixing one type error will make both expressions well-typed.

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/add-pair-example.pdf}
    \caption{
With traditional type error messages, one expression is blamed for causing the type error. This expression may not always match the programmers' intention. 
% \todo{this figure unreadable...}
    }
    \label{fig:single-candidate}
\end{figure}

@@

Deduction steps are motivated by the lack of explaining why the program has type errors. When the compiler rejects a program, it dumps the internal state of type checking.  This information may be a result of complex computing but this process is not reported by compilers. For a typical type error shown in Fig. \ref{fig:ghc-error-example}, clearly, the evidence for the type error is gathered from the previous two declarations. These have to be rediscovered by the programmers again, using harder and less-sound methods.

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/ghc-error-example.png}
    \caption{
        }
    \label{fig:ghc-error-example}
\end{figure}

@@

Programmers use their mouse and keyboard shortcuts to increment or decrement the step number or jump to any step. Typically, a programmer resolves  type errors by navigating through all the deduction steps and verifying whether each explanation aligns with the intended meaning of the program. Eventually, she will find a step that does not match her intention, and the type error can be fixed by modifying one of the two outlined error locations.

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/step-interface.pdf}
    \caption{
Deduction steps if they are shown all at once (left). In practice, steps are shown one at a time. Programmers increment or decrement the step number using the step control bar (Fig. \ref{fig:anatomy}-G) or by directly clicking on a step button (Fig. \ref{fig:anatomy}-F). To increment or decrement the deduction step can be intuitively thought of as moving the position of the \textit{split point} where the blue and orange highlights divide (right).
        }
    \label{fig:step-interface}
\end{figure}

Internally, deduction steps are different ways to group the error locations into two different highlighted colors. If all the highlights are placed in a list-like data structure, each increment of the step numbers moves the end of one color and the start of the other forward (Fig. \ref{fig:step-interface} right).

@@

Nielson pointed out \cite{jakob_nielsen_usability_1993}  that the two most important issues in designing for usability are understanding the users' tasks and the differences in users. From analyzing how users use \chameleon{}, we realized that the ideal debugging interface should adapt to the specific programmer and programming task. It is common that a programmer wants the debugger to just "show the answer", and it is also possible a programmer wants to dive deeper into the problem domain and search for the optimal solution. To accommodate the need to customize the level of information density and granularity of control, \chameleon{} provides three modes: basic, balanced, and advanced. Programmers can switch between modes by clicking on the mode switching toggles (Fig. \ref{fig:anatomy}-C). The features accessible from different modes are shown in table~\ref{tab:chameleon-features}.

\begin{table}
    \centering
    \begin{tabular}{ l l  }
     Mode & Features \\ \hline
     Basic Mode & Type Compare Tool \\ \hline
     Balanced mode & Type Compare Tool \\
     & Candidate Expression Card \\  \hline
     Advanced mode & Type Compare Tool \\
     & Candidate Expression Card \\
     & Deduction Steps \\
    \end{tabular}
    \caption{\chameleon{} modes and features}
    \label {tab:chameleon-features}
\end{table}

>>>>

\section{Motivation}
The design requirements of \chameleon{} are motivated by limitations of traditional type errors, as documented in a number of studies (e.g.~\cite{yang_improved_2000}), but which we illustrate here with a few motivating examples. 
%These limitations are sourced from the authors' frustration in teaching and writing Haskell. This list of shortcomings of traditional type errors is also strengthened by multiple studies pursuing improvement\cite{yang_improved_2000}.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth,trim=0mm 35mm 0mm 0mm]{images/add-pair-example.pdf}
    \caption{
With traditional type error messages, one expression is blamed for causing the type error. This expression may not always match the programmers' intention. 
    }
    \label{fig:motivation-example}
\end{figure}
\subsubsection{\textbf{Traditional type errors show only limited location}}
Haack and Wells~\cite{haack_type_2004} noted that ``\textit{Identifying only one node or subtree of the program as the error location makes it difficult for programmers to understand type errors. To choose the correct place to fix a type error, the programmer must find all the other program points that participate in the error.}''  In the example of Fig \ref{fig:motivation-example}, the type errors can be fixed in multiple locations. For instance  replacing \texttt{['0'..'9']} on line 1 with \texttt{[0..9]}, or  replacing \texttt{fst x} and \texttt{snd x} on line 2 with \texttt {read (fst x)} and \texttt{read (snd  x)}. In the type error message, only the \texttt{addPair} expression on line 4 was blamed.  In this small example the whole context is visible, but the problem is worse for the programmer in large programs where the lines of code contributing to the type error are far apart in the source code.

\subsubsection{\textbf{Traditional type errors are biased}}
In the example of Fig \ref{fig:motivation-example}, the error message arbitrarily focuses the users' attention on only \texttt{addPair}, while ignoring that the literals in the definition of \texttt{u} may be incorrect. %Technically, this results from the side effect of different unification orders of the internal type-checking technique and has no bearing on what the programmers think and expect. 
Another related form of bias is that traditional type errors are often framed as conflicts between \texttt{Expected type} and \texttt{Actual type}. This framing is standard practice in most typed languages. However, what is the \texttt{expected} and what is \texttt{actual} are a side effect of different unification orders rather than the intention of the programmer. In both of these cases, the error message may lead programmers to falsely believe the validity of parts of code and wrongly accuse others.

\subsubsection{\textbf{Traditional type errors give poor explanations}}
When the compiler rejects a program, the internal state of type checking is the result of complex computation. But the details of this process are hard to explain to users and usually not reported by compilers. For the typical type error shown in Fig.~\ref{fig:motivation-example}, the evidence for the type error is gathered from the previous two declarations. These have to be rediscovered by the programmer, using less rigourous methods.

\subsection{Design Goals of \chameleon{}}
Based on the limitations of traditional type errors, we give the following design requirements for \chameleon{}:

\noindent\textbf{Show} all the possible locations where the type error happened or could have happened.

\noindent\textbf{Explain} type errors avoiding jargon and internal constructs of the type checker.

\noindent\textbf{Do not presume} which expression is to blame for the type error based on the order of computation or which possible type for an expression is 'actual' or 'expected'

\end{itemize}

\section{Chameleon IDE} \label{chameleon}
\begin{figure*}
    \centering
    \includegraphics[width=\textwidth, trim=0mm 10mm 0mm 0mm]{images/atonomy.pdf}
    \caption{
        \textbf{The anatomy of \chameleon{}.}
        %\chameleon{} consists of two panes: an editor pane on the left and a
        %debugging pane on the right. 
        The editor pane (left) is similar to a
        traditional code editor. Fragments of source code may have a highlight
        color (A). Additionally, an explanation layer (B) displays in the editor
        window, if deduction steps are enabled. The debugging window contains three blocks. 
        An error statement and configuration block, containing
        An error statement (D), and, optionally,
        a list of candidate expression cards (E), a list of deduction steps (F), 
        and a control bar (G) to increment/decrement deduction step. 
        The conflicting types
        block shows two alternative types (H) inferred by the type inference
        engine. A relevant type information block to show additional information (I)
        that may help understand type errors.
    }
    \label{fig:anatomy}
\end{figure*}

\chameleon{} comprises two parts: a type inference engine and a novel interactive debugging interface. This separation of concerns allows us to (1) adapt \chameleon{} into multiple front ends, such as IDE extensions, desktop applications, and web applications; and (2) reuse the debugging interface for different programming languages by replacing the back-end type inference engine. The debugging interface is designed from the ground up; the type inference engine is a re-implementation of the original Chameleon with several novel improvements, as described in Section \ref{sec:typeinferenceengine}.

@@

@@

The \chameleon{} debugging interface provides three main features to visualize the relation between error statement and locations in code and to explore the explanation of each error location given by the type inference engine. %Further, programmers  can turn on and off features to suit their preferences and debugging needs.

\paragraph{Type compare tool} \label{sub:type-compare}
% In designing \chameleon{}, we want to argue against the practice of reporting type errors as the conflicts between \texttt{Expected type} and \texttt{Actual type}. Technically, the two alternatives are merely the side effect of different unification orders of the internal type-checking technique and have no bearing on what the programmers think and expect.

The type compare tool shows conflicting types in different colors, each associated with one or more error locations highlighted in a matching colour (Fig.~\ref{fig:compare}).  
%The type error must be fixed by modifying at least one of these locations. 
%Error locations are highlighted with the matching color of the conflicting types. The type compare tool is useful to quickly bisect the type error. 
If the programmers know the expression's intended type (they usually do), they will be able to eliminate half of the possible locations. 
%To make this bisecting effect more pronounced, a user can hover over one of the possible types and only show the relevant locations that contribute to that type. 
A hover interaction over one of the possible types facilitates such bisection, causing only the relevant locations that contribute to that type to be highlighted. 
%This is a convenient way to put the error \emph{under the spotlight}.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth, trim=0mm 10mm 0mm 0mm]{images/intro-compare.pdf}
    \caption{
        \textbf{\chameleon{} minimal output (with candidate expression cards and deduction step disabled)}. \chameleon{} identified the conflicting types for the function \texttt{a} and associated the relevant locations with each type. In addition, \chameleon{} leaves out the right-hand side of the declaration '\texttt{0}' and '\texttt{1}' because they are unrelated.
}
    \label{fig:compare}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\linewidth, trim=0mm 10mm 0mm 0mm]{images/intro-candidate.pdf}
    \caption{
        \textbf{\chameleon{} with only candidate expression cards enabled.}
        Indicates the type error can
        occur in the definition of \texttt{x} or \texttt{y}.
    }
    \label{fig:expression}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\linewidth, trim=0mm 10mm 0mm 0mm]{images/intro-deduction.pdf}
    \caption{
        \textbf{\chameleon{} with both candidate expression cards and deduction steps
        enabled.}
        \chameleon{} explains the type error in four steps. In the screenshot, the active step is step 2, where \chameleon{} shows that the expression \texttt{x} and \texttt{y} should have the same type. 
    }
    \label{fig:deduction}
\end{figure}

@@

% We propose candidate expression cards to make it clear to the user that there are multiple ways to resolve a type conflict. By contrast,  standard compiler error messages such as those of GHC arbitrarily focus the users' attention on only one candidate expression, e.g.\ ``\textit{Couldn't match expected type ‘Char’ with actual type ‘Bool’ in expression x}" where x is the candidate expression. In practice, this candidate expression often does not match programmers' intention (Fig. \ref{fig:single-candidate}).  
A candidate expression is an expression that can be inferred to have two conflicting types. 
When a type error is detected, \chameleon{} provides a list of all candidate expressions and programmers are free to choose the problem to resolve by clicking on one candidate expression card. In the example shown in Fig. \ref{fig:expression}, \texttt{x} and \texttt{y} are both candidate expressions. Fixing either type error will make both expressions well-typed.

@@

% Deduction steps are motivated by the lack of explaining why the program has type errors. When the compiler rejects a program, it dumps the internal state of type checking.  This information may be a result of complex computing but this process is not reported by compilers. For a typical type error shown in Fig. \ref{fig:ghc-error-example}, clearly, the evidence for the type error is gathered from the previous two declarations. These have to be rediscovered by the programmers again, using harder and less-sound methods.

% \begin{figure}
%     \centering
%     \includegraphics[width=\linewidth]{images/ghc-error-example.png}
%     \caption{
%         }
%     \label{fig:ghc-error-example}
% \end{figure}

@@

Programmers can use mouse and keyboard shortcuts to increment or decrement the step number or jump to any step. Typically, a programmer resolves  type errors by navigating through all the deduction steps and verifying whether each explanation aligns with the intended meaning of the program. Eventually, she will find a step that does not match her intention, and the type error can be fixed by modifying one of the two outlined error locations.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth, trim=0mm 10mm 0mm 0mm]{images/step-interface.pdf}
    \caption{
Deduction steps if they are shown all at once (left). In practice, steps are shown one at a time. Programmers increment or decrement the step number using the step control bar (Fig. \ref{fig:anatomy}-G) or by directly clicking on a step button (Fig. \ref{fig:anatomy}-F). To increment or decrement the deduction step can be intuitively thought of as moving the position of the \textit{split point} where the blue and orange highlights divide (right).
        }
    \label{fig:step-interface}
\end{figure}

Internally, deduction steps are different ways to group the error locations into two different highlighted colors. 
If all the highlights are placed in a list-like data structure, each increment of the step numbers moves the end of one color and the start of the other forward (Fig.~\ref{fig:step-interface} right).

@@

Nielson pointed out  that the two most important issues in designing for usability are understanding the users' tasks and the differences in users \cite{jakob_nielsen_usability_1993}. From analyzing how users use \chameleon{}, we realized that the ideal debugging interface should adapt to the specific programmer and programming task. It is common that a programmer wants the debugger to just ``show the answer", and it is also possible a programmer wants to dive deeper into the problem domain and search for the optimal solution. To accommodate the need to customize the level of information density and granularity of control, \chameleon{} provides three modes: basic, balanced, and advanced. Programmers can switch between modes by clicking on the mode switching toggles (Fig. \ref{fig:anatomy}-C). The features accessible from different modes are summarised in the following table:%shown in table~\ref{tab:chameleon-features}.

% \begin{table}
%     \centering
\begin{scriptsize}
\begin{center}
    \begin{tabular}{ l l  }
     \textit{Mode} & \textit{Features} \\ \hline
     Basic Mode & Type Compare Tool \\ \hline
     Balanced mode & Type Compare Tool \\
     & Candidate Expression Card \\  \hline
     Advanced mode & Type Compare Tool \\
     & Candidate Expression Card \\
     & Deduction Steps \\
    \end{tabular}
    \end{center}
\end{scriptsize}
%     \caption{\chameleon{} modes and features}
%     \label {tab:chameleon-features}
% \end{table}

Chameleon was originally a command-line tool developed in the early 2000s to improve type error reporting %and introduced a design for interactive type debugging 
for the Haskell programming language.
Unlike traditional type errors produced by the Glasgow Haskell Compiler (GHC)~\cite{ben_gamari_home_2022}, which uses a Hindley–Milner type inference system, Chameleon infers types using constraint solving. In Chameleon, Constraints are generated from the source code based on typing rules. In addition, each constraint is labeled with the location (line and column numbers) where it is generated. This set of constraints is consistent if the program is well-typed and inconsistent otherwise. When a type error occurs, an efficient algorithm is used to derive a minimal subset of the constraints that still contain inconsistencies. This subset is called a Minimal Unsatisfiable Subset (MUS). From this, Chameleon can report a list of locations, using the labels of constraints that are in the MUS. Stuckey et al.~\cite{stuckey_interactive_2003} showed that program locations linked to the constraints from an MUS are all relevant to the type error and must include the cause of the error. In this paper, we will refer to these locations as \textit{error locations}.

Despite successfully borrowing the underlying ideas, we could not reuse the original implementation of Chameleon since the project language standard and libraries used were out of date. 
Our \chameleon{} implementation extends the original Chameleon approach in a number of ways.
%In addition to what Chameleon can do, a few new type inference features were added in the \chameleon{} implementation.

Using only constraints from the MUS is sufficient to locate the type error, but to recover types from type errors we need  constraints from parts of the program that are irrelevant to the type error.  For instance, consider an ill-typed 2-tuple where two possible types can be assigned: \texttt{(Int, Int)} and \texttt{(Int, String)}. The types reconstructed from Chameleon may be \texttt{(a, Int)} and \texttt{(a, String)}. Although the recovered types are consistent, they introduce a new type variable \texttt{a} making the error message harder to understand.  To solve this issue in \chameleon{}, we first expand the MUS to a maximally satisfiable subset. We do so by removing one constraint from the MUS to get a consistent set, and then repeatedly adding constraints from the global constraint store while maintaining consistency. This process stops when no more constraints can be added. The resulting subset, while not helpful to inform error locations, will produce the most concrete conflicting types, see Fig.~\ref{fig:compare-to-original}.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth, trim=0mm 6mm 0mm 0mm]{images/compare-to-original.pdf}
    \caption{
Reporting the same type error, Chameleon uses
\texttt{Int -> a} and \texttt{Char -> a}, while \chameleon{} uses the 
concrete types \texttt{Int -> Bool} and \texttt{Char -> Bool}.
    }
    \label{fig:compare-to-original}
\end{figure}

In addition, \chameleon{} provides support for type explanation. Similar to the type explanation system  in ~\cite{jun_explaining_2002},  \chameleon{} is able to produce a human-readable explanation, but for type errors. This is achieved by annotating nodes in the abstract syntax tree with constraints and the type of rules applied. We generate an inference history from constraints and accompanying annotations.

% \begin{lstlisting} [
%     language=Haskell, caption={
%     A simple program that is ill-typed. This program generates two constraints from line 1 and one constraint from line 2.
% }, label={listing:ifelse}
% ]
% if a then b else c
% a = "True"
% \end{lstlisting}
\begin{listing}[!ht]
\begin{minted}{haskell}
    if a then b else c
    a = "True"
\end{minted}
\vspace{-3mm}
\caption{A simple program that is ill-typed. It generates two constraints from line 1 and one constraint from line 2. }
\label{listing:1}
\end{listing}
For instance, for the program in the listing \ref{listing:1}, \chameleon{}  generates the following constraints and labels (in brackets) $T_a = Bool$ (if condition), $T_b = T_c$ (if branches), $T_a= String$  (definition). Clearly, as $T_a$ can not unify with both \textit{Bool} and \textit{String}, this program is not well typed. \chameleon{} can construct a human-readable explanation from the MUS, for example, \texttt{a} has type \texttt{Bool} because \texttt{a} is the condition of an if statement; however, \texttt{a} has type \texttt{String} because \texttt{a} is defined as the string literal \texttt{"True"}. This low-level explanation is used in a high-level user interface described in section \ref{sub:deduction-steps} to help programmers reason about the validity of error locations.

### 2022-12-08 sections/conclusion.tex
More Tim's edits

We present \chameleon{}, a type debugging tool for the Haskell programming language. \chameleon{} facilitates a constraint-based type inference engine that provides accurate and comprehensive type error reports. \chameleon{} also features an interactive debugging interface with three different debugging modes, enabling programmers to navigate deep type relations and triangulate conflicting statements in source code. Four experiments with human participants were conducted to evaluate the tool's design with programmers with varied experiences. We found that, particularly for more complex tasks, \chameleon{} helped programmers to fix type errors more quickly than traditional text-based error messages.  Further, programmers actively using \chameleon{} interactive features are shown to fix type errors faster than simply reading the type error output.

>>>>

We present \chameleon{}, a type debugging tool for the Haskell programming language. Its constraint-based type inference engine provides accurate and comprehensive type error reports. \chameleon{} also features an interactive debugging interface with three different debugging modes, enabling programmers to navigate deep type relations and triangulate conflicting statements in source code. Our studies evaluated the tool's design with programmers. We found that, particularly for more complex tasks, \chameleon{} helped programmers to fix type errors more quickly than traditional text-based error messages.  Further, programmers actively using \chameleon{} interactive features are shown to fix type errors faster than simply reading the type error output.
